Prefetching SPARQL query cacher

### 2016-07-27 dissertation/cache-proxy.tex
note forward proxy and HERMES
smaller corrections, thanks to @kasei.

This chapter describes an effort to create a proxy that can cache the
results of not only SPARQL queries, but also the results of individual
triple patterns. It will asynchronously analyse executed queries, and
may prefetch the results of certain triple patterns into the cache.

@@

The present system may be deployed at any of these levels. However,
since the database and the reverse proxy in front of it may have
detailed knowledge of the data, e.g. a complete data profile with
statistics to optimise join order, these levels have much in common
with a conventional database cache, which is extensively discussed in
the literature.

@@

It is further interesting to note the contrast to the HERMES
\cite{adali1996query} system's concept of invariants discussed in
Section~\ref{sec:relcache}: a SPARQL query cache be oblivious
to domain knowledge in a CDN deployment and therefore shouldn't assume any invariants,
except in the possible but unlikely case where the remote server
declares an infinite freshness lifetime for the result of a certain
graph pattern.

Figure~\ref{fig:messaging} illustrates the case where HTTP messages
are being passed, where the cache is assumed to reside on a
intermediate caching proxy in the Internet, i.e. a CDN. The figure may
be viewed as having time on the $x$-axis, with the arrows pointing
down being requests and the arrows point up being responses. In this
example, the client first makes a request (request \circled{1}), which
is passed through the proxy because it has not yet cached any
responses. The response (response \circled{1}) is then also
transmitted directly from the SPARQL Endpoint at the database back to
the client. The entire response may be cached on the proxy in case it
can be reused in its entirety. However, the main point is that the
request will be analysed by the proxy in parallel to being sent to the
server, and another request \circled{2} for a triple pattern fragment
that the analyser thinks will enable the proxy to assist the server
the best for future requests, is sent. This analysis is discussed in
Section~\ref{sec:analpre}. The response \circled{2} is not sent back
to the client, but will enter the cache. Lets say that the analyser
was successful and that the cache can be used to answer request
\circled{3}. In that case, the proxy will send a rewritten query for
the rest of the data it needs to be able to answer the query to the origin
server, indicated by the dashed arrows. This may be a SPARQL query, a
TPF request or a combination thereof, depending on what the planner
finds least expensive. The proxy will then evaluate the entire query
and send the final result back as response \circled{3}. In addition,
the analyser may have identified another Triple Pattern Fragment that
it takes as likely to be useful for future requests, and sends a request and
caches the response \circled{4}. At some point, it may be able to
answer the entire query using only cached results, as illustrated by
request and response \circled{5}.

@@

The Attean framework was born from the experiences the PerlRDF
community gained with \pmodule{RDF::Trine}, the Perl counterpart to
the more well-known Jena framework, and the opportunities presented by
the introduction of traits \cite{traits}, or roles as they are known,
to the Perl world.

@@

Perl is an interpreted language and is not known to run fast. Instead,
the focus is on developer efficiency. However, our motivation for
Attean, expressed in \cite{williamspushing}, was to enable the use
of faster implementations by using roles to simplify code reuse.

@@

Also of importance are the APIs to compose stores. Stores in Attean
can be triplestores or quadstores, they can be mutable,
bulkupdateable, can enable the use of ETags or modification times for
conditional requests (as defined in RFC7232 \cite{rfc7234}, and makes
possible further extensions. Each of these primitives is represented
by a role that implements default functionality or simply requires
such functionality to be required. For instance, a quadstore is
required by \pmodule{Attean::API::QuadStore} to implement the
\pcode{get\_quads} method, which can take an RDF term as one or more
of the arguments and variables for the rest, and should return an
iterator over the quads that were matched. Based on an implementation
of the \pcode{get\_quads} method, the \pmodule{Attean::API::QuadStore}
role provides default implementations of methods \pcode{count\_quads},
\pcode{count\_quads\_estimate}, \pcode{get\_graphs} and
\pcode{size}. These methods may be overridden when composing a
class. Composing other roles will require further methods to be
implemented.

Stores can be rather diverse, so in addition, Attean provides another
abstraction layer, called a Model. Models will consistently have
quads, in the case where underlying store is a triple store, a graph
name will be set, or it may wrap several triple stores with a graph
name for each. In addition, Models provide higher level methods, like
for example a \pcode{subject} method to list all subjects, and
corresponding for other terms. Thus, upper layers of an application
should use the Model. We have also chosen to implement significant
parts of the query planning in the Model.

When a query is parsed, algebra objects are generated. Based on the
algebra tree, models may generate plans for any algebra object of
their choosing by implementing the \pmodule{Attean::API::CostPlanner}
role. This amounts to implementing \pcode{plans\_for\_algebra} and
\pcode{cost\_for\_plan} methods. For a given algebra, Attean's query
planner will trust that the model's \pcode{plans\_for\_algebra} is
better than its own by default, a choice that was justified in
\cite{williamspushing}. Additionally, Attean has a simpler mechanism
for single triple or quad patterns: By default, triple or quad
patterns are evaluated by calling the store's \pcode{get\_triples} or
\pcode{get\_quads}, but this may be augmented by adding a wrapper
around an \pcode{access\_plans} method used by the query planner to
produce further plans.

@@

The cache itself is assumed to only cache triples where at least one
term is bound. I have used the Redis data structure store for the
cache as well. This choice is somewhat arbitrary, and has not proved
very successful. The cache will store the results as an array if two
RDF terms are bound in the triple pattern, or a hash if only one term
is bound. The strings put into the cache are serialized N-Triples
strings.

>>>>

This chapter describes an effort to create a proxy that can cache the
results of not only complete SPARQL queries, but also the results of individual
triple patterns. It will asynchronously analyse executed queries, and
may prefetch the results of certain triple patterns into the cache.

@@

The present system may be deployed at any of these levels (and similar
figures could be drawn). However,
since the database and the reverse proxy in front of it may have
detailed knowledge of the data, e.g. a complete data profile with
statistics to optimise join order, these levels have much in common
with a conventional database cache, which is extensively discussed in
the literature.

@@

It is further interesting to note the contrast to the HERMES
\cite{adali1996query} system's concept of invariants discussed in
Section~\ref{sec:relcache}: a SPARQL query cache be oblivious
to domain knowledge in a CDN deployment and therefore shouldn't assume any invariants,
except in the possible but unlikely case where the remote server
declares an infinite freshness lifetime for the result of a certain
graph pattern. A forward proxy, on the other hand, may have a deeper
understanding on the needs of the client, and could possibly implement
invariants. The Attean framework accommodates for this.

Figure~\ref{fig:messaging} illustrates the case where HTTP messages
are being passed, where the cache is assumed to reside on a
intermediate caching proxy in the Internet, i.e. a CDN. The figure
may be viewed as having time on the $x$-axis, with the arrows pointing
down being requests and the arrows point up being responses. In this
example, the client first makes a request (request \circled{1}), which
is passed through the proxy because it has not yet cached any
responses. The response (response \circled{1}) is then also
transmitted directly from the SPARQL Endpoint at the database back to
the client. The entire response may be cached on the proxy in case it
can be reused in its entirety. However, the main point is that the
request will be analysed by the proxy in parallel to being sent to the
server, and another request \circled{2} for a triple pattern fragment
that the analyser thinks will enable the proxy to assist the server
the best for future requests, is sent. This analysis is discussed in
Section~\ref{sec:analpre}. The response \circled{2} is not sent back
to the client, but will enter the cache. Lets say that the analyser
was successful and that the cache can be used to answer request
\circled{3}. In that case, the proxy will send rewritten queries for
the rest of the data it needs to be able to answer the query to the
origin server, indicated by the dashed arrows. This may be zero or more SPARQL
query, TPF requests or a combination thereof, depending on what the
planner finds least expensive. The proxy will then evaluate the entire
query and send the final result back as response \circled{3}. In
addition, the analyser may have identified another Triple Pattern
Fragment that it takes as likely to be useful for future requests, and
sends a request and caches the response \circled{4}. At some point, it
may be able to answer the entire query using only cached results, as
illustrated by request and response \circled{5}.

@@

The Attean framework was born from the experiences the PerlRDF
community gained with \pmodule{RDF::Trine}, the Perl counterpart to
the more well-known Jena framework, and the opportunities presented by
the introduction of traits \cite{traits}, or roles as they are known
to the Perl world.

@@

Perl is an interpreted language and is not focused on speed. Instead,
the focus is on developer efficiency. However, our motivation for
Attean, expressed in \cite{williamspushing}, was to enable the use
of faster implementations by using roles. It also serves to simplify 
code reuse.

@@

Also of importance are the APIs to compose stores. Stores in Attean
can be triplestores or quadstores, they can be mutable,
bulkupdateable, can enable the use of ETags or modification times for
conditional requests (as defined in RFC7232 \cite{rfc7234}), and makes
possible further extensions. Each of these primitives is represented
by a role that implements default functionality or simply requires
such functionality to be required. For instance, a quadstore is
required by \pmodule{Attean::API::QuadStore} to implement the
\pcode{get\_quads} method, which can take an RDF term as one or more
of the arguments and variables for the rest, and should return an
iterator over the quads that were matched. Based on an implementation
of the \pcode{get\_quads} method, the \pmodule{Attean::API::QuadStore}
role provides default implementations of methods \pcode{count\_quads},
\pcode{count\_quads\_estimate}, \pcode{get\_graphs} and
\pcode{size}. These methods may be overridden with more efficient
implementations when composing a class. Composing other roles will
require further methods to be implemented.

Stores can be rather diverse, so in addition, Attean provides another
abstraction layer, called a Model. Models consistently operate over
quads, in the case where underlying store is a triple store, a graph
name will be set, or it may wrap several triple stores with a graph
name for each. In addition, Models provide higher level methods,
for example a \pcode{subject} method to list all subjects, and
corresponding for other terms. Thus, upper layers of an application
should use the Model abstraction. We have also chosen to implement significant
parts of the query planning in the Model.

When a query is parsed, algebra objects are generated. Based on the
algebra tree, models may generate plans for any algebra object of
their choosing by implementing the \pmodule{Attean::API::CostPlanner}
role. This amounts to implementing \pcode{plans\_for\_algebra} and
\pcode{cost\_for\_plan} methods. For a given algebra, Attean's query
planner will trust that the model's \pcode{plans\_for\_algebra} is
better than its own by default, a choice that was justified in
\cite{williamspushing}. Additionally, Attean has a simpler mechanism
for single quad patterns: By default, quad patterns are evaluated by
calling the Model's \pcode{get\_quads}, but this may be augmented by
adding a wrapper around an \pcode{access\_plans} method used by the
query planner to produce further plans.

@@

The cache itself is assumed to only cache the results of triple
patterns where at least one term is bound. 
I have used the Redis data structure store for the
cache as well. This choice is somewhat arbitrary, and has not proved
very successful. The cache will store the results as an array if two
RDF terms are bound in the triple pattern, or a hash if only one term
is bound. The strings put into the cache are serialized N-Triples
strings.
